{memes}: an R interface to the MEME Suite

### 2020-09-21 manuscript.Rmd
starting Results section

```{r setup}
library(dremeR)
library(magrittr)
```

@@

>>>>

```{r setup}
library(memes)
library(magrittr)
```

@@

Here we briefly highlight each of `memes` current features for analyzing
ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
tool, and a complete example using `memes` to analyze ChIP-seq data can be found
in the `memes` vignette and the package website
([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
example, we reanalyze recent work examining the causal relationship between the
binding of the transcription factor E93 to changes in chromatin accessibility
during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
utilize ChIP-seq peak calls for E93 which have been annotated according to the
change in chromatin accessibility observed before & after E93 binding. These
data are an ideomatic example of range-based genomic data (E93 ChIP peaks)
containing additional groupings (the chromatin accessibility response following
DNA binding) whose membership may be influenced by differential motif content.
We show how `memes` syntax allows complex analysis designs, how `memes`
utilities enable deep interrogation of results, and how `memes` flexible data
structures empower users to integrate the `memes` workflow with tools offered by
other R/Bioconductor packages.

### RETURN HERE
### NOTE TO SELF:
 - This is too much of a tutorial & vignette.
 - move this text to the vignette since there isn't any.
 - Instead, highlight some of the interface useage like so:

To run dreme, where `peaks` are chip peaks below with a metadata column `e93_respnose`:
```{r}
seq <- peaks %>% split(mcols(.)$response) %>% get_sequence
runDreme(seq, control = "static")
```

----------------
```{r, include=FALSE}
drake::loadd(chip_results)
```

We begin with a `GenomicRanges` object of E93 ChIP-peak summits +/- 100 bp with
a metadata column indicating the change in chromatin accessibility associated
with E93 binding at that region. 
```{r}
chip_results %>% head
```

### Denovo motif discovery using DREME

One approach to identify *de-novo* motifs within ChIP peaks is to use DREME,
which performs discriminative motif detection in an input set of sequences
relative to a control set \cite{dreme}. To search for *de-novo* motifs within
E93 bindings sites using DREME, we first use the `get_sequence()` function which
returns the DNA sequence within genomic coordinates.
```{r}
genome <- BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3

all_seq <- chip_results %>% 
  get_sequence(genome)
```

The output is stored as a `DNAStringSet`, shown below.
```{r}
head(all_seq, 3)
```

The `runDreme` command is used to begin a motif search using DREME. Although it
is possible to use any number of input sequences, runtimes quickly become
infeasible as sequences exceed 100000 total bp. This can be overcome in a
variety of ways: users can rank peaks by ChIP-seq signal, or some other
quantitative value, and use a subset selected by rank, or in the absence of any
quantitative measure, users can randomly sample their sequences to use as input.
For simplicity in this example, we subset our data to 1000 randomly selected
peaks. Setting `control = "shuffle"` uses DREME's built-in dinucleotide
shuffling algorithm to produce a control set of seqeunces.

```{r}
set.seed(123) # setting a random seed ensures sampling produces a reproducible subsample
sequence_subset <- sample(all_seq, 1000)
all_dreme_res <- runDreme(sequence_subset, control = "shuffle", nmotifs = 3)
```

```{r}
all_dreme_res
```

### Preparing a motif database for searching

Transcription factors often bind with sequence specificity to regulate activity
of nearby genomic elements. Therefore, comparison of *de-novo* motifs with known
DNA binding motifs for transcription factors can be a key step in identifying
regulators of genomic elements.

The TomTom tool is used to compare unknown motifs against a list of known motifs
to identify matches. By passing the results from DREME into TomTom and searching
within a database of *Drosophila* transcription factor motifs, we can identify
candidate transcription factors which may also bind at E93 ChIP peaks. A major
limitation of this approach is that transcription factors containing similar
families of DNA binding domain often possess highly similar motifs, making it
difficult to identify the "true" binding factor of a discovered motif.
Additionally, when searching for matches against a motif database, TomTom must
account for multiple testing, therefore using a larger than necessary motif
database can produce a large multiple testing penalty, limiting sensititity of
detection. One way to overcome these limitations is to limit the transcription
factor motif database to include only motifs for transcription factors expressed
in the sample of interest. Accounting for transcription factor expression during
motif analysis has been demonstrated to  increase the probability of identifying
biologically relevant transcription factor candidates \cite{hector franco's STAR paper, others?}.

The `memes` universalmotif data.frame structure can be used to integrate
expression data with a motif database. First, we import a pre-filtered list of
genes expressed in a timecourse of **Drosophila** wing development.

```{r}
rna_counts <- read.csv("data/wing_expressed_genes.csv", stringsAsFactors = FALSE)
```
```{r, echo = FALSE}
head(rna_counts)
```

Next we import a database of known *Drosophila* transcription factor motifs
generated from the Fly Factor Survey in `universalmotif` format, and convert to a data.frame format using `as_universalmotif_dataframe()` \cite{fly factor}.
```{r}
flyFactor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
  as_universalmotif_dataframe()
```
```{r, echo = FALSE}
head(flyFactor, 5)
```

```{r}
tomtom_results <- runTomTom(all_dreme_res, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
```
Here we visualize the top 3 matches from TomTom for each *de-novo* motif using
`view_tomtom_hits()`. Results are plotted in descending order according to match
score. The results of this function can be integrated into additional R graphics
libraries, such as `patchwork` or `cowplot` to generate publication-quality
figures \cite{patchwork, cowplot}.
```{r, fig.height=5, fig.width=12}
tomtom_results %>% 
  view_tomtom_hits(3) %>% 
  patchwork::wrap_plots(., nrow = 1, tag_level = "new") +
    patchwork::plot_annotation(tag_levels = "A")
```
## Replace above chunks with sections below 
(rstudio crashed without saving)
--------------
Next we import a database of known *Drosophila* transcription factor motifs
generated from the Fly Factor Survey in `universalmotif` format, and convert to a data.frame format using `as_universalmotif_dataframe()` \cite{fly factor}.
```{r}
fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
				  as_universalmotif_dataframe()
	```
	```{r, echo = FALSE}
	head(fly_factor, 5)
	```

The motif database can be subsequently filtered to include only those entries
which match an expressed gene. After filtering, the database is converted back
to `universalmotif` format using `as_universalmotif`.
```{r}
# in this database, altname stores the gene symbol
fly_factor_expressed <- fly_factor %>% 
				  dplyr::filter(altname %in% expressed_genes$symbol) %>% 
					  as_universalmotif()
		```

Passing the results object from *de-novo* discovery with `runDreme` and the
filtered motif database to `runTomTom`
```{r}
tomtom_results <- runTomTom(all_dreme_res, database = fly_factor_expressed, dist = "ed")
tomtom_results <- runTomTom(all_dreme_res, database = fly_factor, dist = "ed")
```

----------------

 - prepare database
  - filter FPKM counts, note this needs to be chosen empirically and is not always a good idea,
  - we do this for demonstration purposes only
 - Denovo discovery in sensitivity categories (static as background)
  - TOMTOM matching, view_match_motifs(3)
    - try patchwork to combine them into figure
 - AME Across binding groups
  - reduce redundant hits
  - plot heatmap
 - FIMO within sensitivity
  - rederive PWM for E93 within sites
   - note differences in specific bases
